---
title: "TDC 2020 Syllabus"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: 
    margin_references: true
    toc: true
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
library(knitr)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

# Program Goals 
The Applied Data Analytics (ADA) programs are targeted at government agency staff, intended to provide training in core data analytics techniques by working on specific projects using real-world administrative microdata. 

The goal of the TANF Data Collaborative program is to help professionals at TANF and related human services agencies develop key data science skills using a hands on approach to working with TANF, Unemployment Insurance Wage Records and employer level data. The ADA program will culminate in a final project based on our [Project Template ](project_template.html){target="_blank"}.

# Program Structure
The program consists of one self-paced module in which participants learn the basics of R and SQL.  The second module is a two week online program. Participants from pilot states are expected to view topic videos before each class. Over the course of 7 days across 2 weeks, participants will spend one hour a day in an interactive classroom session and one hour a day on project work with a facilitator. The third module is a three day onsite program. Participants will have access to the video lectures and notebooks ahead of time, and are expected to watch the videos and work through the notebooks on their own time. There are “checkpoints” throughout each notebook, which are short exercises for participants to complete; checkpoints will be reviewed during interactive classroom sessions.

## Important Dates
Module 1 (completed independently): May 11, 2020 - June 05, 2020

ADRF Orientation (30 min virtual meeting between ADA facilitator and individual teams): June 1-4, 2020 

Module 2: June 08, 2020 - June 19, 2020

Module 3: September 21, 2020 - September 23, 2020

## Estimated Time Use
Based on our schedule, we estimated the hours participants can expect to spend in the course. For independent work, we encourage participants to spend as much time as they need with the material.

```{r echo = FALSE}
time_use_df <- data.frame(matrix(NA, nrow = 6, ncol = 4))

rownames(time_use_df) <- c("Week of May 11", "Week of May 18","Week of May 25","Week of June 1", "Week of June 8","Week of June 15")
colnames(time_use_df) <- c("Activity","Independent work","Instruction","Total")

time_use_df[1:3,1] <- "Module 1"
time_use_df[1:3,2] <- rep("3",3)
time_use_df[1:3,3] <- rep("1",3)
time_use_df[1:3,4] <- rep("4",3)

time_use_df[4,1] <- "ADRF Workspace Orientation; Independent review of video lectures and notebooks"
time_use_df[4,2] <- "3"
time_use_df[4,3] <- "<1"
time_use_df[4,4] <- "3-4"

time_use_df[5:6,1] <- "Module 2"
time_use_df[5,2:4] <- c("4","4","8")
time_use_df[6,2:4] <- c("3","3","6")

kable(time_use_df, caption = "Hrs/Week")

```


# Curriculum Overview
The curriculum is designed to develop the data science skills necessary to work with confidential data, as shown in Figure 1. 

Figure 1: Graphic from  Big Data and Social Science

The goal of the course is to provide a framework for describing the relationship between the labor market faced by TANF recipients and their earnings and employment outcomes. With the final projects, participants will compare earnings and employment outcomes, looking at outcomes of a 2016 cohort and comparing to outcomes in 2009, during the Great Recession. `r margin_note('see the [Table Creation Handout](table_creation_handout.html){target="_blank"} ')` This analysis can be used to shed light on the current crisis. 


We will address the following questions, through lessons and a final project:   

1.  What are the earnings and employment outcomes of TANF recipients?
1.  What are the characteristics of employers likely to hire TANF recipients?
1.  How are employer characteristics related to the labor market outcomes of TANF recipients.
1.  How can you use these results to inform your agency response to the current jobs crisis?

_Policy Relevance: The TANF program “is designed to help needy families achieve self-sufficiency.” State agencies administering this federal program have an interest in understanding how TANF participation can support long-term, stable employment to achieve these goals. Administrative data on TANF participation joined with administrative data on earnings and employers can be used to describe the labor market outcomes of TANF recipients._

# Module 1: Pre-Course Training
There are four separate videos and Notebooks which are structured to be completed online over four consecutive weeks and take no more than 4 hours per week.. Each week includes a one-hour discussion session via Zoom for any questions on the material. The pre-course material will cover basic SQL and R, the latter conducted in Jupyter Notebooks.

# ADRF Orientation
ADA facilitators will join teams via video-conference for up to 30 minutes to ensure that all agreements have been signed and security training completed, to give participants a tour of the ADRF and to get participants connected to the database. At the end of this session, participants should be comfortable navigating within the ADRF, and will have created and tested a database connection.

Learning Objectives:   
1.  Be comfortable navigating the ADRF environment, and be aware of where course materials are located.    
1.  Understand their responsibilities for handling data in the ADRF as per data use agreements and security training    
1.  Open Jupyter notebooks and re-familiarize with the environment.   
1.  Create and test database connection from within Jupyter notebook   

# Module 2: June 8 – June 18, 2020

_Week 1 - Week of June 8_    
Pre Day 1: Three Chapin Hall videos introducing the Chapin Hall data model  
Day 1:  Database Management: Tables and Schema  
Day 2: Data Exploration: Going from basic statistics to developing analytical frame to generating measures   
Day 3: [Applications of Data Exploration](data_exploration_handout.html){target="_blank"}   
Day 4:  Basics of Data Visualization   

_Week 2 - Week of June 15_  
Day 5: Application of Data Visualization   
Day 6: Record Linkage/Data Viz part 2  
Day 7: Application of Record Linkage/Data Viz part 2  
Day 8: Initial Presentations   



**Day 1: Database Management: Tables and Schema**

At the end of this module, participants will be familiar with the tables used in the course notebooks and project template.  They will be introduced to the project template. Students should have already watched a series of three 10-minute lectures by Emily Wiegand from Chapin Hall and will have the opportunity to ask questions about the lectures and engage in discussion with course leaders. Participants will be given discussion questions ahead of time.


Learning Objectives:   
1.  Understand the motivation for the course: describing labor markets in the state and the labor market for TANF exiters
1.  Understand the class structure and the way in which the template will inform their project work and the interaction between notebooks and projects
1.  Become familiar with the primary schemas and tables that will be used in the course.
1.  Understand the basic structure of the primary tables and how they will yield the outcome measures, which will be discussed on Day 2 in video lectures and during class time discussion, and created in the notebooks on Day 3.

**Schedule:**

11 am - 12pm:
Video call for:   

    - Q+A on Emily’s videos   
    - Discussion of Julia’s discussion questions   
    - Discussion of project template

12 - 1pm Project Work:  

    - review of project template  
    - discussion - tie in to final project  

1 - 2 pm Project Work:  

    - review of project template
    - discussion - tie int o final project

Use Slack for additional questions; video-based office hours available upon request


**Day 2: Data Exploration:**

At the end of this module, participants will have learned about the analytical framework - the interaction of TANF recipients and firms in the labor market.  They will learn how to describe the labor market (firm characteristics) and develop standard measures of earnings and employment outcomes, as well as earning trajectories. They will learn what questions to consider when developing an analysis using such measures. The lesson will range from covering basic statistics to developing an analytical framework for creating employment measures. Participants will be introduced to the cohorts of TANF leavers that will be used for analysis.  This material will lay the groundwork for the hands-on data exploration tasks covered in Day 3.

_Policy Relevance: Understanding employment trajectories of TANF recipients in connection with firms is the primary focus of this course, and through a lecture and discussion session, participants will understand the primary measures of earnings and employment outcomes, particularly whether to look at annual earnings, earnings from dominant employers or all employers, etc. They will also learn how to characterize employers._

Learning Objectives:  
1.  Become familiar with common measures of earnings and employment outcomes. 
1.  Begin to think about which measures are most important to their home state’s pilot project. 
1.  Develop measures of employer characteristics
1.  Learn about the cohorts of TANF leavers that they will be working with on Day 3.   
1.  Be prepared for the hands-on Data Exploration notebook which will be reviewed tomorrow. 


**Schedule:**
11 am - 12pm:
Video call for:

    - Q+A on Data Exploration videos   
    - discussion of Julia’s discussion questions  
    - foreshadow data exploration notebook which will be discussed tomorrow  

12 - 1pm Project Work - 4 teams

1 - 2 pm Project Work - 4 teams

Use slack for additional questions; video-based office hours available upon request


**Day 3: Applications of Data Exploration**
Participants will learn to query, filter and aggregate the data and create basic descriptive statistics of employment and earnings of a pre-defined cohort. Using pre-written code on a cohort  of TANF exiters, participants will calculate employment and earnings measures for the cohort post-exit.  They will also describe the characteristics of employers that hire them. In the notebook checkpoints, they will modify code to generate the same measures in a comparison cohort. 

Learning Objectives:  

1.  Learn to query the database using SQL.
1.  Learn to use the main tables in TANF and UI wage record data.
1.  Gain skills in creating basic descriptives of the datasets, aggregating data frames.
1.  Learn to calculate earnings and employment outcome measures (given a pre-defined cohort) which were discussed on Day 2. 

**Schedule:**
11 am - 12pm:
Video call for:  

    - Q+A on Data Exploration notebook  
    - review of checkpoints  
    - get participants to explain their answers to the checkpoint exercises  

12 - 1pm Project Work - 4 teams

1 - 2 pm Project Work - 4 teams

Use slack for additional questions; video-based office hours available upon request


**Day 4: Basics of Data Visualization**
At the end of this module, participants will have learned foundational concepts in data visualization, and be exposed to specific visualizations pertinent to their study of employment trajectories. 

Learning Objectives: 

1.  Learn about best practices in visualization  
1.  Be exposed to important visualizations (e.g the 8-quarter post-exit visualization)


**Schedule:**
11 am - 12pm:
Video call for:  

    - Q+A on Julia’s Data Viz video 
    - discussion of Julia’s discussion questions  
    - foreshadow Data Viz notebook which will be reviewed tomorrow

12 - 1pm Project Work - 4 teams

1 - 2 pm Project Work - 4 teams

Use slack for additional questions; video-based office hours available upon request

**Day 5: Applications of Data Visualization**

At the end of this module, participants will have hands-on experience creating basic tables and plots using the statistics and measures calculated  earlier in the week.  Participants will use summary statistics and measures created on **Day 3: Applications of Data Exploration** and the concepts learned on **Day 4: Basics of Data Visualization** to create visualizations including distribution of different employment and earnings across a cohort, time series plots showing patterns over time, box plots, Sankey plots. 
 
_Policy Relevance: With the summary statistics and measures created earlier in the week, participants will create visuals conveying employment patterns across the cohorts of interest after exit, e.g., after 1-, 5- and 8- quarters post-exit._

Learning Objectives:  

1.  Become familiar with ggplot2 syntax   
1.  Generate at plots conveying employment outcomes with varying inputs   
1.  Generate visualizations for post-exit employment  

**Schedule:**
11 am - 12pm:
Video call for:  

    - Q+A on Data Viz notebook 
    - review of checkpoints 
    - get participants to explain their answers to the checkpoint exercises

12 - 1pm Project Work - 4 teams

1 - 2 pm Project Work - 4 teams

Use slack for additional questions; video-based office hours available upon request

**Day 6: Record Linkage**

At the end of this module, participants will have been exposed to various approaches to record linkage and have a deeper understanding of how deterministic record linkage (joining) will be used with these data. This material will lay the groundwork for the hands-on linkage tasks covered in **Day 7: Applications of Record Linkage.**

_Policy Relevance: Joining TANF and UI Wage Record data gives analysts the opportunity to look at whether individuals interact with (e.g, re-enroll) in the TANF program post-exit. _

Learning Objectives:  

1.  Be exposed to approaches to record linkage (deterministic, probabilistic, CS/ML approaches)
1.  Understand why deterministic linkage is used for these data
1.  Be prepared to create the post-exit pattern calculation and visualization, specifically by incorporating ‘return to TANF’ outcomes in the viz.

**Schedule:**
11 am - 12pm:
Video call for:  

    - Q+A on Julia’s Record Linkage video 
    - discussion of julia’s discussion questions
    - foreshadow record linkage notebook which will be reviewed tomorrow

12 - 1pm Project Work - 4 teams

1 - 2 pm Project Work - 4 teams

Use slack for additional questions; video-based office hours available upon request

**Day 7: Applications of Record Linkage/Viz Part 2**

At the end of this module, participants will be able to join data from TANF and UI Wage record data, and look at patterns in return to TANF. 

_Policy Relevance: In joining these datasets, participants will see the power/expanse of available research questions that the joined data provides. They will further understand their populations by examining missingness (individuals on welfare who do not have employment records) and comparing lifespans and overlap of time of participation in welfare and the labor market._

Learning Objectives:
1. Become comfortable joining UI Wage Record Data and TANF datasets
2. Update post-exit visualizations created in Day 5: Applications of Visualization by adding squares to represent return to TANF.
