---
title: "Table Creation Handout"
author: "Coleridge"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: 
    margin_references: true
    toc: true
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
library(knitr)
library(yaml)
library(tidyverse)
library(magrittr)
setwd("/Users/sophierand/Desktop/ada-tdc-2020")
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)

code_snips_file <-"/Users/sophierand/Desktop/ada-tdc-2020/code_snips.yml"
code_snips_list <- yaml::read_yaml(code_snips_file, fileEncoding = "UTF-8")

code_snips <- as.data.frame(matrix(unlist(code_snips_list), nrow=length(code_snips_list), byrow=TRUE))
colnames(code_snips) <- names(code_snips_list[[1]])

```


# Introduction

```{marginfigure, echo=TRUE}
See `Table Creation.ipynb`  
```

This document covers the specifications of cohort tables and the SQL queries used to generate them. 

# TANF Tables
```{marginfigure, echo=TRUE}
There is one row per `ssn`; for individuals who had more than one spell ending in the exit period, the most recent spell is in the table (e.g., the spell with the latest report month where `tanf_end` == `TRUE`.)
```
TANF tables follow the naming convention: `ada_tdc_2020.tanf_<year>q<quarter>`. Each of these tables contains individuals who were primary recipients of TANF and had a spell ending in the year-quarter in the table name. The unit of analysis is `person-spell`. 

Two tables were created in the TANF data: `ada_tdc_2020.tanf_2009q1` and `ada_tdc_2020.tanf_2016q4` which contain spells for primary TANF recipients who ended a spell in Q1 2009 or Q4 2016, respectively. `r margin_note('See data dictionary *chapin_hall_model.xlsx*')` 

## Form of Table

```{marginfigure, echo=TRUE}
`tanf_total_months` is sum of `tanf_spell_months` across all spells that the individual had.
```

```{r, echo = F}
df <- data.frame(matrix(nrow = 4, ncol = 7))
colnames(df)<-c("ssn","month","tanf_start","tanf_end","tanf_spell_months","tanf_total_months","county")

df[,1] <- c("ssnA","ssnB","ssnC","ssnD")
df[,2] <- c("200901","200902","200901","200903")
df[,3] <- c(FALSE, TRUE, FALSE, FALSE)
df[,4] <- c(TRUE, TRUE, TRUE, TRUE)
df[,5] <- c(4,1,7,13)
df[,6] <- c(10,1,18,13)
df[,7] <- c(097, 052,095,173)

kable(df,align=rep('c', 7))
```

## Specifications   

_Inclusion Criteria_   
- Spell ends ("exit") in a given quarter (2016 Q4 and 2009 Q1)  
- Recipient is a primary recipient of TANF   

_Characteristics_   
- Earliest Time Stamp: quarter of exit   
- Latest Time Stamp: 8 quarters post-exit


## SQL Create Table Query
```{marginfigure, echo=TRUE}
`tanf_end = TRUE` indicates end of a spell    
```

```{marginfigure, echo=TRUE}
`relat == 1` indicates that recipient is a primary recipient
```

_TANF Spells for recipients in 2009 Q1 Cohort_
```yaml
`r code_snips$code[code_snips$title=="ada_tdc_2020.tanf_2009q1"]`

```

_TANF Spells for recipients in 2016 Q4 Cohort_
```yaml
`r code_snips$code[code_snips$title=="ada_tdc_2020.tanf_2016q4"]`


```


These queries were generated using the `sprintf` function  
_TANF Spells for recipients in Cohort (flexible query)_
```yaml

`r code_snips$code[code_snips$title=="ada_tdc_2020.tanf_sprintf_query_inputs"]`


`r code_snips$code[code_snips$title=="ada_tdc_2020.tanf_sprintf_query"]`


```

# Wage Record Tables

Wage tables follow the naming convention: `ada_tdc_2020.tanf_wage_<year><quarter>`. Each of these contains all wage records for the SSNs in the TANF cohort tables. This is a longitudinal file and records span from the quarter of exit until 8 quarters post exit. The tables were created using the UI Wage Record data: `ada_tdc_2020.tanf_wage_2016q4` and
`ada_tdc_2020.tanf_wage_2009q1` which contain job and wage data. The unit of analysis is a `person-job`.


## Specifications

_Inclusion Criteria_     
- Contains ssns that were in the corresponding cohort     
- Wage record data for exit-quarter and 8 quarters post-exit   
 
_Characteristics_    
- Earliest Time Stamp: quarter of exit     
- Latest Time Stamp: 8 quarters post-exit   

## Form of Table


```{r, echo = F}
wage_df <- data.frame(matrix(nrow = 4, ncol = 6))
colnames(wage_df)<-c("ssn","year","quarter","uiacct","wages","job_yr_q")

wage_df$ssn <- c("ssnA","ssnB","ssnC","ssnD")
wage_df$year<-c(2010,2010,2010,2011)
wage_df$quarter <-  c(2,3,1,1)
wage_df$uiacct <- c("S5GBU48PyzI0","xT5jPKk7pBn2","XsvUJdMmKikw", "CPRrceZIwBGl")
wage_df$wages <- c(1291,200,3866,1928)
wage_df$job_yr_q<- c("2010-04-01","2010-07-01","2010-01-01","2011-01-01")


kable(wage_df,align=rep('c', 6))
```

## SQL Create Table Query

_Wages for Individuals in 2009 Q1 Cohort_

```yaml
`r code_snips$code[code_snips$title=="ada_tdc_2020.tanf_wage_2009q1"]`

```

_Wages for Individuals in 2016 Q4 Cohort_
```yaml
`r code_snips$code[code_snips$title=="ada_tdc_2020.tanf_wage_2016q4"]`

```

# Employer-Wage Record Tables

Employer tables are base tables for generating employer characteristics like 

All Employment:   

    - total payroll  
    - average payroll  
    - total employment  
    - earnings at the 25th percentile  
    - earnings at the 75th percentile  

Stable Employment:   

    - total full quarter payroll  
    - average full quarter payroll  
    - total full quarter employment  
    
    
```{marginfigure, echo=TRUE}
Firms who employ TANF leavers are determined by filtering for firms where an `ssn` from the TANF cohorts appear in their wage records. 
```
Tables will be created for all firms, in two time periods. In Data Exploration, we will filter this table to look at characteristics of firms that employ TANF leavers.

## Specification   

`r margin_note('confirm that the time specifications on this are correct')`

_Inclusion Criteria_    

    - Contains wage data for all firms   
    - Wage records in Exit quarter + 2 quarters post-exit  

_Characteristics_     

    - Earliest Time Stamp: quarter of exit   
    - Latest Time Stamp: 2 quarters post-exit

## Form of Table


```{r, echo = F}
wage_df <- data.frame(matrix(nrow = 4, ncol = 10))
colnames(wage_df)<-c("ssn","year","quarter","uiacct","fein","wages","naics_3_digit", "cnty","egr","id")

wage_df$ssn <- c("ssnA","ssnB","ssnC","ssnD")
wage_df$year<-c(2010,2010,2010,2011)
wage_df$quarter <-  c(2,3,1,1)
wage_df$uiacct <- c("S5GBU48PyzI0","xT5jPKk7pBn2","XsvUJdMmKikw", "CPRrceZIwBGl")
wage_df$fein <- rep('fein', 4)
wage_df$wages <- c(1291,200,3866,1928)
wage_df$naics_3_digit<-c(447,772,624,724)
wage_df$cnty <- rep('a county code', 4)
wage_df$egr <- rep("?",4)
wage_df$id <- rep("?",4)

kable(wage_df,align=rep('c', 10))
```

## SQL Create Table Query

_All Employers with wages in 2016Q4 exit quarter + 2 quarters_

```yaml
`r code_snips$code[code_snips$title=="all_empl_wage_2016q4"]`

```

_All Employers with wages in 2009Q1 exit quarter + 2 quarters_
```yaml
`r code_snips$code[code_snips$title=="all_empl_wage_2009q1"]`

```


# Employer Characteristics Tables

## Form of Table
```{marginfigure, echo=TRUE}
`naics_3_digit` is the industry identifier for business. 

**CLayton** - is it confusing that there is a year/quarter field since theres only 1 in this dataset?
```

```{r, echo = F, fig.fullwidth = TRUE}
empl_df <- data.frame(matrix(nrow = 4, ncol = 13))
colnames(empl_df)<-c("fein","naics_3_digit", "year","quarter","job_yr_q","tot_employment",
                     "tot_earnings","avg_earnings","bottom_25_earnings","top_25_earnings",
                     "full_quarter_employment","full_quarter_earnings","full_quarter_avg_earnings")

empl_df$fein <- rep('fein', 4)
empl_df$naics_3_digit<-c(447,772,624,724)
empl_df$year<-c(2010,2010,2010,2011)
empl_df$quarter <-  c(2,3,1,1)
empl_df$job_yr_q<- c("2010-04-01","2010-07-01","2010-01-01","2011-01-01")
empl_df$tot_employment <- NA
empl_df$tot_earnings <- NA
empl_df$avg_earnings <- NA
empl_df$bottom_25_earnings <- NA
empl_df$top_25_earnings <- NA
empl_df$full_quarter_employment <- NA
empl_df$full_quarter_earnings <- NA
empl_df$full_quarter_avg_earnings <- NA


empl_char_df <- data.frame(matrix(nrow = 13, ncol = 2))
colnames(empl_char_df) <- c("column","description")
empl_char_df$column<-c("fein","naics_3_digit", "year","quarter","job_yr_q","tot_employment",
                     "tot_earnings","avg_earnings","bottom_25_earnings","top_25_earnings",
                     "full_quarter_employment","full_quarter_earnings","full_quarter_avg_earnings")
# empl_char_df$description <- c("employer identifier")
empl_char_df <- empl_char_df %>% 
  mutate(description = case_when(
    column =="tot_employment" ~ "count(ssn) in reference quarter",
    column == "avg_earnings" ~ "sum(wages)/count(ssn) in reference quarter",
    column == "tot_earnings" ~ "sum(wages) in reference quarter",
    column == "bottom_25_earnings" ~ "earnings, all individuals, at the 25th percentile",
    column == "top_25_earnings" ~ "earnings, all individuals, at the 75th percentile"
  ))

kable(empl_char_df)
```


## Specifications

The base table for employment characteristics are wages from firms in the quarter of exit and 2 quarters post-exit. Tables are `all_empl_wage_2016q4` and `all_empl_wage_2009q1`.

`r margin_note('confirm that the time specifications on this are correct')`

_Inclusion Criteria_    

    - Wage records for firms, starting in quarter post-exit   
    - Contains wage data for all firms     
    - For measures of stable employment, wage records for individuals who were employed in exit quarter + 2 quarters post-exit    
    
_Characteristics_  

The below employment and earnings measures are calculated for wages in the quarter after exit.   

    - total payroll  - sum(wages) in reference quarter  
    - average payroll  - sum(wages)/count(ssn) in reference quarter  
    - total employment  - count(ssn) in reference quarter  
    - earnings at the 25th percentile    
    - earnings at the 75th percentile    

The below measures, calculated on _stably employed individuals_, defined as individuals employed for 3 consecutives quarters by the same firm, are calculated with a reference quarter of the quarter after exit (2009 Q2 and 2017 Q1).
```{marginfigure, echo=TRUE}
For more on these earnings and employment measures, see the Bureau of Labor Statistics' [Quarterly Workforce Indicators 101](https://qwiexplorer.ces.census.gov/static/explore.html#x=0&g=0)
```

    - total full quarter payroll  
    - average full quarter payroll  
    - total full quarter employment  
    

## SQL Create/Alter/Update Table Query

Employer characteristics are developed with SQL `CREATE`, `ALTER`, and `UPDATE` statements. 

_2016 Employers_
```{marginfigure, echo=TRUE}
The wage records are filtered for the quarter post-exit (Q1 2017), which will be the _reference quarter_ for the full-quarter (stable) employment measures.
```
```yaml
`r code_snips$code[code_snips$title=="all_empl_char_2016q4"]`

```

Alter table to add columns for stable employment indicators

```yaml
`r code_snips$code[code_snips$title=="alter_all_empl_char_2016q4"]`

```

Create base table employing a `CROSS JOIN` that will be used to calculate the stable employment (full-quarter employment) measures and populated into the `all_empl_char_2016q4` columns we just created. 


```{marginfigure, echo=TRUE}
A `CROSS JOIN` is employed to create a longitudinally linked file. The CROSS JOIN is done by _crossing_ the table with itself three times, joining wages from the reference quarter to wages from the quarter before and after the reference. This is done bcause full-quarter employment requires at least 3 months of employment at hte same firm. See [Appendix](#Appendix) for more on the `CROSS JOIN`.
```

```yaml
`r code_snips$code[code_snips$title=="create_stable_empl_base_2016q4"]`

```

Using that base table, we calculate the full-quarter employment metrics and update.

```yaml
`r code_snips$code[code_snips$title=="update_all_empl_char_2016q4_from_base"]`

```


_2009 Employers_
```yaml
`r code_snips$code[code_snips$title=="all_empl_char_2009q1"]`

```

Alter table to add columns for stable employment indicators

```yaml
`r code_snips$code[code_snips$title=="alter_all_empl_char_2009q1"]`

```

Create base table employing a `CROSS JOIN` that will be used to calculate the stable employment (full-quarter employment) measures and populated into the `all_empl_char_2016q4` columns we just created. 

```yaml
`r code_snips$code[code_snips$title=="create_stable_empl_base_2009q1"]`

```

Using that base table, we calculate the full-quarter employment metrics and update.

```yaml
`r code_snips$code[code_snips$title=="update_all_empl_char_2009q1_from_base"]`

```

# Appendix
## Full Quarter Employment

FQE is a proxy for stable employment - it represents employment of an individual for at least three quarters by the same employer. 


$$fullQuarterEmployent = num_{ssn}$$   
where $$ssn \in wagerecords_{q1,q2,q3}  $$   
and $$employer_{q1}=employer_{q2}=employer_{q3}$$


The table needed to calculate employer characteristics, e.g. the full-quarter employment measures, arises from a `CROSS JOIN` of wage record tables. The cross join is done on three keys: `ssn`, `fein` and `job_yr_q`, to ensure that the table only includes individuals who were employed in _consecutive_ quarters, at the _same job_ (denoted by `fein`).

## Specifications
_Inclusion Criteria_   
- Inidividuals who had three consecutive quarters of employment at the same firm
- Looking at stably employed inidividuals with a reference quarter as the first quarter after exit of our TANF cohorts

The form of the table is effective equal to that of the wage tables. However, only those individuals who were employed for three consective quarters, beginning with the exit quarter, are considered _stably employed_ and contribute to the aggregate statistics. 

_Form of the table created with a `CROSS-JOIN`, the base table for employment outcomes_
```{r, echo = F, fig.fullwidth = TRUE}
cross_join_df <- data.frame(matrix(nrow = 3, ncol = 9))
colnames(cross_join_df)<-c("ssn","year","quarter","uiacct","wages",
                           "fein","job_yr_q", "lag_q","lead_q")

cross_join_df$ssn <- rep('ssn', 4)
cross_join_df$year<-c(2010,2010,2010,2011)
cross_join_df$quarter <-  c(2,3,1,1)
cross_join_df$uiacct <- c("S5GBU48PyzI0","xT5jPKk7pBn2","XsvUJdMmKikw", "CPRrceZIwBGl")
cross_join_df$wages <- c(1291,200,3866,1928)
cross_join_df$fein <- rep('fein', 4)
cross_join_df$job_yr_q<- c("2010-04-01","2010-04-01","2010-04-01")
cross_join_df$lag_q <- c("2010-04-01","2010-04-01","2010-04-01")
cross_join_df$lead_q <- c("2010-04-01","2010-04-01","2010-04-01")

  
kable(cross_join_df)
```

_All employment outcomes_

_Stable employment outcomes_